Potential and limits of site-specific substitution model inference.

### 2020-01-18 manuscript/ms.tex
typos and streamlining. Move supp figures to supplement

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

@@

\begin{abstract}
Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
Some sites evolve close to neutral, while others are highly conserved, allow only specific states, or only change in concert with other sites.
Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
We find that the interative approximate maximum likelihood scheme uses information in the data efficiently and results in accurate  estimates of site specific preferences from large data sets with moderately diverged sequences.
We find that ignoring site specific preferences during branch length estimation of phylogenetic trees -- an assumption of all popular phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
Site-specific preferences estimated from large HIV \emph{pol} alignments show good concordance with intra-host estimates and highlight the inability of estimating deep divergence times of HIV and SIVs using molecular clock approaches due to saturation.
\end{abstract}

@@

\section*{Introduction}
Over time, genome sequences change through mutations and are reshuffeled by recombination.
Modifications to the genomes are filtered by selection for survival such that beneficial variants spread preferentially and those that impair function are purged.
As a result, some parts of genomes change rapidly, while other are strongly conserved.
In addition to variation of the evolutionary rate, different sites in a genome only explore different subsets of the available states.
Some positions in a protein, for example, might only allow for hydrophobic amino acids, while others require acidic side chains.
Patterns of conservation and variation, possibly involving more than one site, are therefore shaped by functional contraints which in turn allows inference of biological function from genetic variation.

@@

Inferring phylogenies is a computationally challenging problem since the number of phylogenies grows super-exponentially with the number of taxa and because the calculation of the likelihood is computationally costly (though linear in the number of taxa).
Due to this computational complexity, the most commonly used substitution models are simple caricatures of biological complexity.
The simplest substitution models assume that all sites and sequence states are equivalent and evolve at the same rate, i.e., they assume an unconstrained non-functional sequence that mutates at random between the different sequence states.
Such simple models are clearly inadequate and ignoring rate heterogeneity tends to result in biased estimates of divergence times or otherwise erroneous results \citep{yang1996among}.
Most commonly used models account for variation in substitution rates among sites and average properties of the substitution process such as transition/transversion bias or more likely substitution between similar amino acids \citep{yang_maximum_1994,FastTree2,nguyen_iq-tree:_2015,stamatakis_raxml_2014}.
To avoid over-fitting, these methods typically don't estimate rates for each site, but treat site-specific rates as random effects that are integrated out (often using discrete approximations of a Gamma distribution \citep{yang1996among}, mixture of multiple uni-model distributions \citep{mayrose2005gamma}, or a small number of fixed rates).

In addition to rate variation, different sites in a protein differ in amino acid they allow.
Recent deep mutational scanning experiments have shown that site specific preferences are mostly conserved between moderately diverged proteins \citep{doud_site-specific_2015}.
Using such experimentally inferred site specific models in phylogenetic inference greatly increases the likelihood of the data \cite{bloom2014experimentally}.
More than two decades ago, \citet{halpern_evolutionary_1998} pointed out that ignoring that equilibrium frequencies vary from one position to another will result in underestimation of branch lengths of a phylogeny -- possibly dramatically when frequencies are heavily skewed.
\citet{hilton_modeling_2018} recently showed that experimentally measured preference not only improve the phylogenetic fit, but also results in longer branch length estimates.
Models with site-specific preferences are also known as mutation-selection balance models \citep{bruno1996modeling,yang2008mutation} reflecting the intuition that equilibrium frequencies are determined by competition of diversifying processes (like mutation) and selection for an optimal function.

While biologically plausible, estimating such models from data exacerbates the over-fitting problem and it rarely attempted in practice.
In the context of the site-specific models this issue has become known as {\em extensive parametrization} or even {\em infinitely many parameters problem} \cite{Rodrigue557,spielman2016extensively}.
With sufficient data, however, site specific parameters can be accurately estimated \citep{tamuri_estimating_2012,spielman2016extensively,scheffler2014validity}.
Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on the fidelity with which site specific model parameters are known.

\begin{figure*}[htb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha1.5}.pdf}
	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations. Colored lines correspond to data generated assuming Gamma distributed rate variation with $\alpha=1.5$, grey lines with $\alpha=3$.
	The estimates are more accurate than naive estimates from the state frequencies in alignment columns. The advantage over naive estimates increases with data set size as seem from different curves with $n\in [100,300,1000,3000]$ sequences per tree.
	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates. Regularization via pseudo-counts reduces over-fitting at low divergence. }
	\label{fig:dressed}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Model
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Efficient inference of site specific substitution models}
Following previous \citet{halpern_evolutionary_1998}, we assume that the substitution matrix $W_{ij}$ between states $j,i$ does not depend on the position in the sequence, while
the overall rates and the equilibrium frequencies depend on the site.
Such a site-specific general time-reversible (GTR)) model, can be parameterized as:
\begin{eqnarray}
Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
\label{eq:Qij}
\end{eqnarray}
where $W_{ij}$ is a symmetric matrix with $W_{ii}=0$ and the second equation ensures conservation of probability.
In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to ensure that the average rate per site is $\mu^{a}$.

Extending the iterative expectation maximization approach by \citet{bruno1996modeling}, we show in the appendix that the iterative update rules
\begin{equation}
\label{eq:update}
	\begin{split}
		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
	\end{split}
\end{equation}
approximately maximize the likelihood of the data.
Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree and $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$
$\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ divided by the total tree length.
To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated. This (i) shows that the update rules are multiplicative and therefore ensure positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
These update rules are an example of non-negative factorization algorithms \citep{lee_algorithms_2001}.

@@

To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
We quantify the accuracy of the inference as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities and compare inference using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_nuc_ratealpha1.5}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_aa_ratealpha1.5}.pdf}
	\caption{{\bf Quantification of errors stemming from tree inference and ancestral reconstruction.}
	Panels A and B show the mean-squared deviation $\chi^2$ of inferred $\hat{\eqp}_i^a$ from the true $\eqp_i^a$ for 4-letter and 20-letter alphabets, respectively.
	At large root-to-tip distances, ancestral reconstruction becomes less and less certain and estimation of $\eqp_i^a$ fails (red lines).
	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch length using the updated models (or using the true tree, yellow/pink).
	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.}
	\label{fig:reconstructed}
\end{figure*}

Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site for estimates of $\eqp_i^a$ from the alignment and using the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
The plot is shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope 1 implies that the accuracy is inversely proportional to the expected number of substitutions.
This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
The data shown in Fig.~\ref{fig:reconstructed} was generated by drawing site-specific rates from a Gamma distributed with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
The scaling behavior $\chi_2\sim 1/\textrm{tree length}$ is much more evident for $\alpha=3$ than for $\alpha=1.5$ as the average accuracy is sensitive to a small number of sites with low rates at which frequencies are estimated poorly.
A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:reconstructed}A).

@@

%%% effect of reconstruction
In a typical scenario, ancestral sequences are unknown and need to be inferred or summed over (marginalized).
To test the influence of tree and ancestral state reconstruction, we built phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model or FastTree \citep{FastTree} using the default 20 category model for simulated nucleotide and amino-acid sequences, respectively.
Fig~\ref{fig:reconstructed} compares different schemes to reconstruct ancestral sequences, infer substitution models, and optimize the tree.
The simplest approach is to take the inferred tree as given, reconstruct the ancestral states using a standard evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^b_{ij}$ and $\tau_i^b$ from this reconstruction.
This naive approach works well up to root-to-tip distances of about 0.3, beyond which estimates deteriorate, see Fig~\ref{fig:reconstructed} (``Reconstructed ancestral sequences'', red line).
Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest improvement in accuracy (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}.
More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site specific model (``Iterative model estimation'', brown line).
The accuracy now continues to improve up to root-to-tip distances of about one.
At this level of divergence, tree reconstruction starts becoming problematic and branch lengths deviate substantially from their true values.
Using the true tree instead of the reconstructed tree leads to continuous improvements of accuracy with increasing levels of divergence (yellow line).
Similar improvements are achieved by optimizing tree branch lengths along with the model.

We found little difference between 4-letter and 20-letter alphabets in the overall accuracy of the estimated model and its dependence on the different approximations, compare Fig.~\ref{fig:reconstructed}A\&B.

\subsection*{Ignoring site specific frequencies results in branch length underestimates}
As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^a}^2$ is increasing sharply with more peaked preferences.
Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent rate \citet{wertheim_purifying_2011}.

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/nuc_length_n1000}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/nuc_depth_n1000}.pdf}
	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates}.
	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length. Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation. Not that inferred site specific models only partially ameliorate underestimation, see main text. Parameters: $n=1000$, $\alpha=1.5$. }
	\label{fig:tree_length}
\end{figure*}

%% dependence on pseudo count
Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted trees (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR=R10 model without site specificity.
Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogeny which are prone to underestimation.
When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
This effect is entirely due to skewed equilibrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^a=q^{-1}$ with the same rate variation.
%% dependence on nucleotide and amino acid skew
The underestimation of branch lengths is less of a problem when using larger alphabets and the true probability are not too heavily skewed, see Fig.~\ref{fig:tree_length_supp}.

@@

Surprisingly, the inferred site specific model only partially rectifies the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
The principle contributor to this deviation are inaccuracies in the rate estimates.
Combining the true rates with inferred preferences reduces the error in branch length estimation (see Fig.~\ref{fig:tree_length}, lines labeled ``(true rates)''.

@@

Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
Deviation in rate estimates result immediately in underestimated branch length, while substantial effect of too flat site specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
If the same preferences are used for every site ($\alpha=1$), however, deviations are substantial.
Deviations in rate and preferences are approximately additive.
These observations are consistent with the finding above using inferred preferences and true substitution rates result in more accurate branch length estimates than inferring both rates and preferences.

@@

\subsection*{Applications to large HIV alignments}
Since its zoonosis, HIV-1 has diversified into several different subtypes that differ from each other at about 20\% of sites in their genome.
For each subtype, thousands of sequences are available, which typically differ by about 10\% \citep{LANL}.
This large sample of moderately diverged sequences should be suited to apply our iterative inference frame work.
We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site specific GTR models at the nucleotide and amino acid level, see Materials and Methods.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
\usepackage{xr}
\usepackage{hyperref}
\externaldocument{supplement}
%\usepackage{cite}

@@

\begin{abstract}
Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
Some sites evolve close to neutrally, while others are highly conserved, allow only specific states, or only change in concert with other sites.
Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and results in accurate  estimates of site-specific preferences from large data sets with moderately diverged sequences.
Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
Site-specific preferences estimated from large HIV \emph{pol} alignments show qualitative concordance with intra-host estimates of fitness costs.
Analysis of site-specific HIV substitution models suggests near saturation of divergence after a few hundred years.
Such saturation can explain the inability to infer deep divergence times of HIV and SIVs using molecular clock approaches and time-dependent rate estimates.
\end{abstract}

@@

\section*{Introduction}
Over time, genome sequences change through mutations and are reshuffled by recombination.
Modifications to the genomes are filtered by selection for survival such that beneficial variants spread preferentially and those that impair function are purged.
As a result, some parts of genomes change rapidly, while other are strongly conserved.
In addition to variation of the evolutionary rate, different sites in a genome only explore different subsets of the available states.
Some positions in a protein, for example, might only allow for hydrophobic amino acids, while others require acidic side chains.
Patterns of conservation and variation, possibly involving more than one site, are therefore shaped by functional constraints which in turn allows inference of biological function from genetic variation.

@@

Inferring phylogenies is a computationally challenging problem since the number of phylogenies grows super-exponentially with the number of taxa and because the calculation of the likelihood is computationally costly (though linear in the number of taxa).
Due to this computational complexity, the most commonly used substitution models are simple caricatures of biological complexity.
The simplest substitution models assume that all sites and sequence states are equivalent and evolve at the same rate, i.e., they assume an unconstrained non-functional sequence that mutates at random between the different sequence states.
Such simple models are clearly inadequate and ignoring rate heterogeneity tends to result in biased estimates of divergence times or otherwise erroneous results \citep{yang1996among}.
Most commonly used models account for variation in substitution rates among sites and average properties of the substitution process such as transition/transversion bias or more likely substitution between similar amino acids \citep{yang_maximum_1994,FastTree2,nguyen_iq-tree:_2015,stamatakis_raxml_2014}.
To avoid over-fitting, these methods typically don't estimate a rate for each site, but treat site-specific rates as random effects that are integrated out (often using discrete approximations of a Gamma distribution \citep{yang1996among}, mixture of multiple unimodal distributions \citep{mayrose2005gamma}, or a small number of fixed rates).

In addition to rate variation, different sites in a protein differ in amino acid they allow.
Recent deep mutational scanning experiments have shown that site-specific preferences are mostly conserved between moderately diverged proteins \citep{doud_site-specific_2015}.
Using such experimentally inferred site-specific models in phylogenetic inference greatly increases the likelihood of the data \cite{bloom2014experimentally}.
More than two decades ago, \citet{halpern_evolutionary_1998} pointed out that ignoring that equilibrium frequencies vary from one position to another will result in underestimation of branch lengths of a phylogeny -- possibly dramatically when frequencies are heavily skewed.
\citet{hilton_modeling_2018} recently showed that experimentally measured preference not only improve the phylogenetic fit, but also results in longer branch length estimates.
Models with site-specific preferences are also known as mutation-selection balance models \citep{bruno1996modeling,yang2008mutation} reflecting the intuition that equilibrium frequencies are determined by competition of diversifying processes (like mutation) and selection for an optimal function.

While biologically plausible, estimating such models from data exacerbates the over-fitting problem and it rarely attempted in practice.
In the context of the site-specific models this issue has become known as {\em extensive parametrization} or even {\em infinitely many parameters problem} \cite{Rodrigue557,spielman2016extensively}.
With sufficient data, however, site-specific parameters can be accurately estimated \citep{tamuri_estimating_2012,spielman2016extensively,scheffler2014validity}.
Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on the fidelity with which site-specific model parameters are known.
We apply this algorithm to large HIV-1 alignments and explore the consequences for phylogenetic inference.

\begin{figure*}[htb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha1.5}.pdf}
	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations. Colored lines correspond to data generated assuming Gamma distributed rate variation with $\alpha=1.5$, gray lines with $\alpha=3$.
	The estimates are more accurate than naive estimates from the state frequencies in alignment columns. The advantage over naive estimates increases with data set size as seem from different curves with $n\in [100,300,1000,3000]$ sequences per tree.
	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates. Regularization via pseudo-counts reduces over-fitting at low divergence.
	Analogous results for alphabets of size $q=20$ are shown in Fig.~\ref{fig:dressed_supp}.
}
	\label{fig:dressed}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Model
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Efficient inference of site-specific substitution models}
Following work by \citet{halpern_evolutionary_1998}, we parameterize a site-specific general time-reversible (GTR) substitution model at site $a$ from state $j$ to $i$ as:
\begin{eqnarray}
Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki} \ .
\label{eq:Qij}
\end{eqnarray}
Here, $\mu^a$ is the substitution rate at site $a$, $\eqp^a_i$ is the equilibrium frequency of state $i$ at site $a$, and $W_{ij}$ is a symmetric substitution matrix that we assume to be the same for all sites.
The second equation ensures conservation of probability.
In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to normalize the frequencies and fix the scale of $W_{ij}$.

Extending the approach by \citet{bruno1996modeling}, we show in Materials and Methods and the supplement that the iterative update rules
\begin{equation}
\label{eq:update}
	\begin{split}
		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
	\end{split}
\end{equation}
approximately maximize the likelihood of the data.
Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree, $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$,
$\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ divided by the total tree length.
To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated.
This (i) shows that the update rules are multiplicative and therefore ensure positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
These update rules are an example of non-negative factorization algorithms \citep{lee_algorithms_2001}.

@@

To assess these sources of error independently, we simulated sequences evolving along trees and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
For each parameter set, we generated two random Yule trees, for each tree generated two random site-specific models, and for each tree/model pair two alignments, resulting in a total of eight data set per parameter combination.
We quantified the accuracy of the inferences as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities, respectively.
We compared inferences using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_nuc_ratealpha1.5}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_aa_ratealpha1.5}.pdf}
	\caption{{\bf Quantification of errors stemming from tree inference and ancestral reconstruction.}
	Panels A and B show the mean-squared deviation $\chi^2$ of inferred $\hat{\eqp}_i^a$ from the true $\eqp_i^a$ for 4-letter and 20-letter alphabets, respectively.
	At large root-to-tip distances, ancestral reconstruction becomes less and less certain and estimation of $\eqp_i^a$ fails (red lines).
	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch length using the updated models (or using the true tree, yellow/pink).
	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.
	}
	\label{fig:reconstructed}
\end{figure*}

Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site.
The graph contains estimates of $\eqp_i^a$ from the alignment and via the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
The data are shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope -1 implies that the accuracy is inversely proportional to the expected number of substitutions.
This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
The data shown in Fig.~\ref{fig:reconstructed} was generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
The scaling behavior $\chi_2\sim 1/\textrm{tree length}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
In case of strong rate variation, the average accuracy is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.
A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:reconstructed}A).

@@

%%% effect of reconstruction
In a typical scenario, ancestral sequences are unknown and need to be inferred or summed over (marginalized).
To test the influence of tree and ancestral state reconstruction, we reconstructed phylogenetic trees using IQ-tree \citep{nguyen_iq-tree:_2015} with a GTR+R10 model (simulated nucleotide sequences) or FastTree \citep{FastTree} using the default 20 category (simulated amino-acid sequences).
Fig~\ref{fig:reconstructed} compares different schemes to reconstruct ancestral sequences, infer substitution models, and optimize the tree.
The simplest approach is to take the inferred tree as given, reconstruct the ancestral states using a standard evolutionary model (e.g.~Jukes-Cantor model) and calculate $n^b_{ij}$ and $\tau_i^b$ from this reconstruction.
This naive approach works well up to root-to-tip distances of about 0.3, beyond which estimates deteriorate, see Fig~\ref{fig:reconstructed} (``Reconstructed ancestral sequences'', red line).
Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest improvement in accuracy (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}).
More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site-specific model (``Iterative model estimation'', brown line).
The accuracy now continues to improve up to root-to-tip distances of about one.
At this level of divergence, tree reconstruction starts becoming problematic and branch lengths deviate substantially from their true values.
Using the true tree instead of the reconstructed tree leads to continuous improvements of accuracy with increasing levels of divergence (yellow line).
Similar improvements are achieved by optimizing tree branch lengths along with the model.
We found little difference between 4-letter and 20-letter alphabets in the overall accuracy of the estimated model and its dependence on the different approximations, compare Fig.~\ref{fig:reconstructed}A\&B.

\subsection*{Ignoring site-specific frequencies results in branch length underestimates}
As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^a}^2$ is increasing sharply with more peaked preferences.
Models that don't account for site-specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent evolutionary rates \citet{wertheim_purifying_2011}.

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/nuc_length_n1000}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/nuc_depth_n1000}.pdf}
	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates}.
	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length. Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation. Not that inferred site-specific models only partially ameliorate underestimation, see main text. Parameters: $n=1000$, $\alpha=1.5$. }
	\label{fig:tree_length}
\end{figure*}

%% dependence on pseudo count
Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted trees (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR+R10 model without site-specificity.
Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogenies which are prone to underestimation.
When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
This effect is entirely due to skewed equilibrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^a=q^{-1}$ with the same rate variation.
%% dependence on nucleotide and amino acid skew
The underestimation of branch lengths is less severe for larger alphabets of if frequencies are not heavily skewed, see Fig.~\ref{fig:tree_length_supp}.

@@

Surprisingly, using the inferred site-specific models only partially rectified the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
The principle contributor to this deviation are inaccuracies in the rate estimates.
Combining the true rates with inferred preferences reduces the error in branch length estimation (see Fig.~\ref{fig:tree_length}, lines labeled ``(true rates)''.

@@

Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
Deviation in rate estimates result immediately in underestimated branch length, while substantial effects of too flat site-specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
If the same preferences are used for every site ($\alpha=1$), however, deviations are substantial.
Deviations in rate and preferences are approximately additive.
These observations are consistent with the finding above that using inferred preferences and true substitution rates result in more accurate branch length estimates than inferring both rates and preferences.

@@

\subsection*{Applications to large HIV alignments}
Since its zoonosis, HIV-1 has diversified into several different subtypes that differ from each other at about 20\% of sites in their genome.
For each subtype, thousands of sequences are available, which typically differ by about 10\% \citep{LANL}.
This large sample of moderately diverged sequences should be suited to apply our iterative inference frame work.
We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site-specific GTR models at the nucleotide and amino acid level, see Materials and Methods.
